---
title: "Going Beyond: Interpreting Survival Models"
---

## Introduction

### Expanding Interpretability to Survival Analysis

We are thrilled to announce the release of `{midr}` 0.6.0.
This update brings the core principles of the MID framework to a highly requested domain: *Survival Analysis*.

With this release, users can now decompose and interpret complex, "black-box" survival models—such as Random Survival Forests (RSF)—just as easily as standard regression and classification models.

Key new features in 0.6.0 include:

- **Native Support for Survival Models**: The `interpret()` function now natively supports some survival models including Cox proportional hazard models (`{survival}`) and random survival forests (`{randomForestSRC}`, `{ranger}`).

- **Time-Dependent Sequence of MID Models**: Because predicted survival probabilities change over time, `{midr}` now calculates and returns an array of time-varying MID surrogate models.

## Example

```{r}
#| message: FALSE
#| warning: FALSE
# required packages
library(survival)
library(randomForestSRC)
library(midr)

library(ggplot2)
library(gridExtra)
library(patchwork)
theme_set(theme_midr())

data(flchain, package = "survival")
```

Here is a look at how `{midr}` 0.6.0 interprets a Random Survival Forest model using the `flchain` dataset.

### Predictive Modeling

```{r}
# predictive modeling: random survival forest
rsf_model <- rfsrc(
  Surv(futime, death) ~ age + sex + sample.yr + kappa + lambda, 
  data = flchain, 
  ntree = 100,
  nodesize = 20,
  na.action = "na.impute"
)

rsf_model
```

### Surrogate Modeling

```{r}
# surrogate modeling
mid_model <- interpret(
  Surv(futime, death) ~ (age + sex + sample.yr + kappa + lambda) ^ 2, 
  data = flchain,
  model = rsf_model,
  lambda = .05,
  na.action = na.pass
)

mid_model
```

### Intercept

Baseline survival probabilities, i.e. the average predicted survival probabilities, are stored as `$intercept`.

Here, we compare the intercepts with the Kaplan-Meier curve.

```{r}
#| fig.height: 3
#| fig.width: 7.25
km_fit <- survfit(
  Surv(futime, death) ~ 1, data = flchain
)

par.midr()

plot(
  km_fit, 
  main = "Kaplan-Meier vs. Average Model Prediction",
#  xlab = "Days from Sample Collection", 
  ylab = "Survival Probability",
  ylim = c(0.5, 1.0)
)

lines(
  x = as.numeric(labels(mid_model)),
  y = mid_model$intercept,
  col = "#FF000080",
  lwd = 2
)
```

### Uninterpreted Variation Ratio

Since `{midr}` fits as many models as the number of time periods of interest, the uninterpreted variation ratio changes with time.

```{r}
#| fig.height: 3
#| fig.width: 7.25
par.midr()

plot(
  x = as.numeric(labels(mid_model)),
  y = mid_model$ratio,
  type = "l",
  main = "Uninterpreted Variation Ratio",
  ylab = "Ratio",
  xlab = "Days from Sample Collection"
)
```

### Feature Importance

```{r}
imp <- mid.importance(mid_model)

p <- ggmid(
  imp,
  theme = "magma",
  terms = rev(head(mid.terms(imp), 10))
) +
  theme(legend.position = "none") +
  coord_flip()

q <- ggmid(
  imp,
  type = "series",
  theme = "Tableau 10",
  terms = mid.terms(imp, interactions = FALSE),
  linewidth = 3/4
) +
  labs(x = "Days from Sample Collection") +
  theme(legend.position = "bottom")

p / q
```

### Feature Effects

```{r}
#| fig.height: 8
#| fig.width: 7.25
plots <- mid.plots(
  mid_model,
  terms = c("age", "lambda", "kappa", "sex"),
  type = "series",
  theme = "mako",
  resolution = 100,
  intercept = TRUE,
  limits = c(-0.05, 1.05)
)

plots[[4]] <- plots[[4]] + labs(x = "Days from Sample Collection")

grid.arrange(nrow = 4, grobs = plots)
```

### Breakdown of Effects

```{r}
brk <- mid.breakdown(mid_model, data = flchain[1 ,])

p <- ggmid(
  brk,
  theme = "cividis",
  terms = rev(head(mid.terms(imp, interactions = FALSE)))
) +
  theme(legend.position = "none") +
  coord_flip()

q <- ggmid(
  brk,
  type = "series",
  theme = "Tableau 10",
  terms = mid.terms(imp, interactions = FALSE),
  linewidth = 3/4
) +
  labs(x = "Days from Sample Collection") +
  theme(legend.position = "bottom")

p / q
```

### Conditional Expectation

```{r}
#| fig.height: 3
#| fig.width: 7.25
con <- mid.conditional(mid_model, variable = "age", data = flchain[100, ])
ggmid(con, theme = "viridis", type = "series") +
  labs(x = "Days from Sample Collection")
```