---
title: "Surrogate Modeling with MID in R"
---

## Introduction

In modern actuarial science, there is an inherent tension between predictive accuracy and model transparency.
While ensemble tree-based models like **Gradient Boosting Machines (GBMs)** frequently outperform traditional **Generalized Linear Models (GLMs)**, their "black-box" nature presents significant hurdles for model governance, regulatory compliance, and price filing.

This notebook demonstrates a solution using **Maximum Interpretation Decomposition (MID)** via the `{midr}` and `{midnight}` packages in R.

### What is MID?

MID is a functional decomposition framework that acts as a high-fidelity surrogate for complex models.
It deconstructs a black-box prediction function $f(\mathbf{x})$ into several interpretable components: intercept $g_\emptyset$, main effects $g_j(x_j)$, and interaction effects $g_{jk}(x_j, x_k)$.
The prediction is represented as the following additive structure:

$$
f(\mathbf{x}) = g_\emptyset + \sum_{j} g_j(x_{j}) + \sum_{j < k} g_{jk}(x_{j},\;x_{k}) + \dots + g_D(\mathbf{x})
$$

To ensure the uniqueness and interpretability of each component, MID imposes vanishing conditional expectation constraints:

$$
\begin{aligned}
\mathbf{E}\left[g_j(X_j)\right] &= 0 \\
\mathbf{E}\left[g_{jk}(X_j, X_k) \mid X_j = x_j\right] &= 0 \quad (\forall x_j) \\
\mathbf{E}\left[g_{jk}(X_j, X_k) \mid X_k = x_k\right] &= 0 \quad (\forall x_k)
\end{aligned}
$$

By replicating a black-box model with this structured approach, we can quantify the "uninterpreted" variance (captured by $g_D(\mathbf{x})$) and derive a representation that captures the superior predictive power of machine learning without sacrificing actuarial clarity.

### Setting Up

We begin by setting up the environment and loading the necessary libraries.

```{r setting, message=FALSE, warning=FALSE}
# data manipulation
library(arrow)
library(dplyr)

# predictive modeling
library(gam)
library(lightgbm)

# surrogate modeling
library(midr)
library(midnight)

# visualization
library(ggplot2)
library(gridExtra)

# load training and testin datasets
train <- read_parquet("../data/train.parquet")
test  <- read_parquet("../data/test.parquet")
```

```{r config, include = FALSE}
# cold execution
COLD_RUN <- FALSE

if (COLD_RUN) {
  set.seed(42)
  train <- train[sample(nrow(train), 50000), ]
  test  <- test[sample(nrow(test), 50000), ]
}
```

A key component of our evaluation is the **Weighted Mean Poisson Deviance** defined as follows.

$$
L(\mathbf{y}, \hat{\mathbf{y}}, \mathbf{w}) = \frac{2 \sum_{i=1}^n w_i \left( y_i \log(y_i/\hat{y}_i) - (y_i - \hat{y}_i) \right)}{\sum_{i=1}^n w_i}
$$

```{r poisson_deviance}
# define loss function
mean_poisson_deviance <- function(
    y_true, y_pred, sample_weight = rep(1, length(y))
  ) {
  stopifnot(all(y_pred > 0))
  resid <- ifelse(y_true > 0, y_true * log(y_true / y_pred), 0)
  resid <- resid - y_true + y_pred
  2 * sum(resid * sample_weight) / sum(sample_weight)
}
```

## The Interpretable Baseline: GAM

We first fit a GAM to establish a transparent benchmark.
Since GAMs are additive by design, they provide a "ground truth" model structure to be recovered by the functional decomposition.

```{r fit_gam}
fit_gam <- gam(
  Frequency ~ s(VehPower) + s(VehAge) + s(DrivAge) + s(LogDensity) +
              VehBrand + VehGas + Region,
  data = train,
  weights = Exposure,
  family = quasipoisson(link = "log")
)

summary(fit_gam)
```

```{r pred_fit_gam}
# evaluate fitted model
pred_fit_gam <- predict(fit_gam, test, type = "response")

deviance <- mean_poisson_deviance(
  y_true = test$Frequency,
  y_pred = pred_fit_gam,
  sample_weight = test$Exposure
)
cat("Mean Poisson Deviance:", deviance)
```

### Surrogate Modeling

We apply the `interpret()` function to the GAM.
This step serves as a sanity check: if MID is effective, it should perfectly replicate the predictive behavior of the original GAM.

```{r mid_gam}
mid_gam <- interpret(
  Frequency ~ VehPower + VehAge + DrivAge + LogDensity +
              VehBrand + VehGas + Region,
  data = train,
  weights = Exposure,
  link = "log",
  model = fit_gam
)

summary(mid_gam)
```

```{r pred_mid_gam}
# evaluate fitted surrogate
pred_mid_gam = predict(mid_gam, test, type = "response")

deviance <- mean_poisson_deviance(
  y_true = test$Frequency,
  y_pred = pred_mid_gam,
  sample_weight = test$Exposure
)
cat("Mean Poisson Deviance:", deviance)
```

### Model Fidelity

To assess model **fidelity**, i.e., how closely the surrogate replicates the black-box, we calculate the **uninterpreted variation ratio** $\mathbf{U}$.

$$
\mathbf{U}(f,g) = \frac{\sum_{i=1}^n (f(x_i) - g(x_i))^2}{\sum_{i=1}^n (f(x_i) - \bar{f})^2}, \quad \text{where } \bar{f} = \frac{1}{n}\sum_{i=1}^n f(x_i)
$$

This metric represents the proportion of the black-box model's variance that is not captured by the additive components of the MID model.
The R-squared score, $\mathbf{R}^2(f,g) = 1 - \mathbf{U}(f,g)$, is a standard measure for this purpose. It is important to note that this $\mathbf{R}^2$ measures the fidelity to the black-box model, not the predictive accuracy relative to the ground truth observations.

In the `{midr}` package, the summary output includes this ratio calculated on the training set.
For models with non-linear links (e.g., Poisson regression), the "working" ratio is computed on the scale of the link function (e.g., $\log$ scale).

To rigorously confirm the model fidelity, it is recommended to evaluate these metrics on a separate testing set.
This ensures that the surrogate model is not just over-fitting the training predictions but has truly captured the underlying functional structure.

```{r }
# calculate R-squared on testing dataset
R2_mid <- weighted.loss(
  x = log(pred_fit_gam),
  y = log(pred_mid_gam),
  w = test$Exposure,
  method = "r2"
)

cat(sprintf("R-squared: %.6f", R2_mid))
```

As shown by the high $\mathbf{R}^2$ score, the MID surrogate achieves near-perfect fidelity.
This level of agreement justifies using the MID components (main effects and interactions) as a reliable lens through which to interpret the original black-box model’s behavior.

### Feature Effects

Visualizing the functional behavior of each component allows for a direct comparison between the MID surrogate's decomposition and the original GAM's structure.

- MID Surrogate

```{r effects_mid_gam}
# main effects of MID surrogate
par.midr(mfrow = c(2, 4))
mid.plots(mid_gam, engine = "graphics", ylab = "Main Effect")
```

- Original GAM

```{r effects_fit_gam}
# feature effects of GAM
par.midr(mfrow = c(2, 4))
termplot(fit_gam)
```

Furthermore, we can visualize the joint effects of feature pairs as 3D prediction surfaces using the S3 method for the `persp()` function.

```{r persp_mid_gam}
par.midr(mar = c(1, 0, 1, 0), mfrow = c(1, 2))
persp(mid_gam, "DrivAge:LogDensity", theta = 45, phi = 40)
persp(mid_gam, "LogDensity:Region", theta = -45, phi = 40)
```

### Effect Importance

Beyond simple plots for feature effects, `{midr}` provides a suite of diagnostic tools.
First, the **Effect Importance** of a term $j$ is defined as the mean absolute contribution of that term across the population:

$$
\text{Importance}_j = \mathbf{E} \left[ | g_j(X_j) | \right] \approx \frac{1}{n} \sum_{i=1}^n | g_j(x_{ij}) |
$$

```{r imp_gam, message = FALSE}
imp_gam <- mid.importance(mid_gam, data = train, max.nrow = 2000)
grid.arrange(
  nrow = 1, widths = c(5, 4),
  ggmid(imp_gam, fill = "steelblue") +
    labs(title = "Effect Importance",
         subtitle = "Average absolute effect per feature"),
  ggmid(imp_gam, type = "beeswarm", theme = "mako@div") +
    labs(title = "",
         subtitle = "Distribution of effect per feature") +
    scale_y_discrete(labels = NULL) +
    theme(legend.position = "none")
)
```

For interaction terms, the importance is similarly calculated using $g_{jk}(X_j, X_k)$.
This metric allows us to rank features by their average influence on the model's predictions.

### Conditional Expectation

Second, we can explore **Individual Conditional Expectations (ICE)**.
In the MID framework, the ICE for a feature $j$ and a specific observation $i$ is the expected value of the prediction as $X_j$ varies, while keeping other features fixed at their observed values $\mathbf{x}_{\setminus j}^{(i)}$:

$$
\text{ICE}_j^{(i)}(x) = g_\emptyset + g_j(x) + \sum_{k \neq j} g_{jk}(x, x_{ik})
$$

```{r ice_gam, message = FALSE}
ice_gam_link <- mid.conditional(mid_gam, type = "link", variable = "DrivAge")
ice_gam <- mid.conditional(mid_gam, variable = "DrivAge")
grid.arrange(
  nrow = 1,
  ggmid(ice_gam_link, var.color = LogDensity) +
    theme(legend.position = "bottom") +
    labs(y = "Linear Predictor",
         title = "Conditional Expectation",
         subtitle = "Change in linear predictor"),
  ggmid(ice_gam, type = "centered", var.color = LogDensity) +
    theme(legend.position = "bottom") +
    labs(y = "Prediction", title = "",
         subtitle = "Centered change in original scale")
)
```

Unlike standard black-box models, MID's low-order structure allows us to compute these expectations efficiently and interpret the variation across curves (the "thickness" of the ICE plot) as a direct consequence of specified interaction terms $g_{jk}$.

### Additive Attribution

Third, we perform instance-level explanation through an Additive Breakdown of the prediction.
For any single observation $\mathbf{x}$, the MID surrogate's prediction $g(\mathbf{x})$ is decomposed into the exact sum of its functional components:

$$
g(\mathbf{x}) = \underbrace{g_\emptyset}_{\text{Intercept}} + \underbrace{\sum_{j} g_j(x_j)}_{\text{Main Effects}} + \underbrace{\sum_{j < k} g_{jk}(x_j, x_k)}_{\text{Interactions}}
$$

```{r}
set.seed(42)
row_ids <- sort(sample(nrow(train), 4))
bd_list <- lapply(
  row_ids,
  function(x) {
    res <- mid.breakdown(mid_gam, train, row = x,
                         format = list(LogDensity = "%.2f"))
    structure(res, row_id = x)
  }
)
bd_plots <- lapply(
  bd_list, function(x) {
    label <- paste0("Breakdown of Row ", attr(x, "row_id"))
    ggmid(x, theme = "shap") +
      labs(x = NULL, subtitle = label) +
      theme(legend.position = "none")
  }
)
grid.arrange(grobs = bd_plots)
```

By visualizing these contributions in a waterfall plot, we can identify which specific risk factors or interaction effects drove the prediction for a particular instance, such as a high-risk policyholder.

## The Black-Box: LightGBM

While GAMs are transparent, GBMs such as **LightGBM** often yield superior predictive power by capturing high-order interactions.
However, this accuracy comes at the cost of being a black box.

```{r}
# hold out validation dataset
valid_idx <- seq_len(floor(nrow(train) * 0.2))

# create datasets for training
dtrain <- lgb.Dataset(
  data.matrix(select(train[-valid_idx, ], -Frequency, -Exposure)),
  label = train$Frequency[-valid_idx],
  weight = train$Exposure[-valid_idx],
  categorical_feature = c("VehBrand", "VehGas", "Region")
)

dvalid <- lgb.Dataset.create.valid(
  dtrain,
  data.matrix(select(train[ valid_idx, ], -Frequency, -Exposure)),
  label = train$Frequency[ valid_idx],
  weight = train$Exposure[ valid_idx]
)

# model parameters
params_lgb <- list(
  objective = "poisson",
  learning_rate = 0.03188002,
  num_leaves = 30,
  reg_lambda = 0.004201069,
  reg_alpha = 0.2523909,
  colsample_bynode = 0.5552524,
  subsample = 0.5938199,
  min_child_samples = 9,
  min_split_gain = 0.3920509,
  poisson_max_delta_step = 0.8039541
)

set.seed(42)
fit_lgb <- lgb.train(
  params = params_lgb,
  data = dtrain,
  nrounds = 1000L,
  valids = list(eval = dvalid),
  early_stopping_round = 50L,
  verbose = 0L
)

summary(fit_lgb)
```

```{r}
# evaluate fitted model
pred_fit_lgb <- predict(
  fit_lgb, data.matrix(select(test, -Frequency, -Exposure))
)

deviance <- mean_poisson_deviance(
  y_true = test$Frequency,
  y_pred = pred_fit_lgb,
  sample_weight = test$Exposure
)
cat("Mean Poisson Deviance:", deviance)
```

### Surrogate Modeling

We now use `{midr}` to replicate the LightGBM model.
By including interaction terms in the model formula, we allow the surrogate to capture the joint relationships that the GBM has learned.
The goal is to approximate the LightGBM function $f_{LGB}(\mathbf{x})$ with our interpretable structure $g(\mathbf{x})$:

$$
f_{LGB}(\mathbf{x}) \approx g(\mathbf{x}) = g_\emptyset + \sum_{j} g_j(x_j) + \sum_{j < k} g_{jk}(x_j, x_k)
$$

```{r}
mid_lgb <- interpret(
  Frequency ~ (VehPower + VehAge + DrivAge + LogDensity +
               VehBrand + VehGas + Region)^2,
  data = train,
  lambda = 0.01,
  weights = Exposure,
  link = "log",
  model = fit_lgb,
  pred.fun = function(model, data) {
    newdata <- data.matrix(select(data, -Frequency, -Exposure))
    predict(model, newdata)
  }
)

summary(mid_lgb)
```

```{r}
# evaluate fitted surrogate
pred_mid_lgb = predict(mid_lgb, test, type = "response")

deviance <- mean_poisson_deviance(
  y_true = test$Frequency,
  y_pred = pred_mid_lgb,
  sample_weight = test$Exposure
)
cat("Mean Poisson Deviance:", deviance)
```

### Model Fidelity

To measure how successfully our surrogate replicates the LightGBM model, we use the R-squared score on the link scale ($\log$ scale). 

```{r}
# calculate R-squared on testing dataset
R2_mid <- weighted.loss(
  x = log(pred_fit_lgb),
  y = log(pred_mid_lgb),
  w = test$Exposure,
  method = "r2"
)

cat(sprintf("R-squared: %.4f", R2_mid))
```

### Feature Effects

```{r}
par.midr(mfrow = c(2, 4))
mid.plots(mid_lgb, engine = "graphics", ylab = "Partial Effect")
```

A key advantage of `{midr}` is its ability to isolate interaction effects $g_{jk}$ from main effects $g_j$.
This is particularly useful to understand the joint impact of two variables (e.g., `Region` and `LogDensity`).

```{r}
grid.arrange(
  nrow = 1, widths = c(3, 2),
  ggmid(mid_lgb, "LogDensity:Region", type = "data",
        data = train[1:1e4, ]) +
    labs(y = NULL, subtitle = "Interaction Effect") +
    theme(legend.position = "bottom"),
  ggmid(mid_lgb, "LogDensity:Region", main.effects = TRUE) +
    labs(y = NULL, subtitle = "Total Effect") +
    scale_y_discrete(labels = NULL) +
    theme(legend.position = "bottom")
)
```

```{r}
par.midr(mar = c(1, 0, 1, 0), mfrow = c(1, 2))
persp(mid_lgb, "DrivAge:LogDensity", theta = 45, phi = 40)
persp(mid_lgb, "LogDensity:Region", theta = -45, phi = 40)

```

### Effect Importance

To rank the influence of each component discovered in the LightGBM model, we calculate the Effect Importance, defined as the average absolute contribution.

```{r, message = FALSE}
imp_lgb <- mid.importance(mid_lgb, data = train, max.nrow = 2000)
grid.arrange(
  nrow = 1, widths = c(4, 3),
  ggmid(imp_lgb, theme = "bluescale@qual", max.nterms = 20) +
    labs(title = "Effect Importance",
         subtitle = "Average absolute effect per feature") +
    theme(legend.position = "none"),
  ggmid(imp_lgb, type = "beeswarm", theme = "mako@div", max.nterms = 20) +
    labs(title = "",
         subtitle = "Distribution of effect per feature") +
    scale_y_discrete(labels = NULL) +
    theme(legend.position = "none")
)
```

### Conditional Expectation

We further explore the model's behavior using the ICE plot.
In the MID framework, the variation in ICE curves for a feature $j$ is explicitly governed by the interaction terms $g_{jk}$ identified from the LightGBM model.

```{r, message = FALSE}
ice_lgb_link <- mid.conditional(mid_lgb, type = "link", variable = "DrivAge")
ice_lgb <- mid.conditional(mid_lgb, variable = "DrivAge")
grid.arrange(
  nrow = 1,
  ggmid(ice_lgb_link, var.color = LogDensity) +
    theme(legend.position = "bottom") +
    labs(y = "Linear Predictor",
         title = "Conditional Expectation",
         subtitle = "Change in linear predictor"),
  ggmid(ice_lgb, type = "centered", var.color = LogDensity) +
    theme(legend.position = "bottom") +
    labs(y = "Prediction", title = "",
         subtitle = "Centered change in original scale")
)
```

### Additive Attribution

Finally, we perform an Additive Breakdown for individual predictions.
This provides an exact allocation of the LightGBM's prediction into the terms of our surrogate model.

```{r}
set.seed(42)
row_ids <- sort(sample(nrow(train), 4))

bd_list <- lapply(
  row_ids,
  function(x) {
    res <- mid.breakdown(mid_lgb, train, row = x,
                         format = c("%.6s", "%.3s, %.3s"))
    structure(res, row_id = x)
  }
)

bd_plots <- lapply(
  bd_list, function(x) {
    label <- paste0("Breakdown of Row ", attr(x, "row_id"))
    ggmid(x, theme = "shap", max.nterms = 10) +
      labs(x = "Linear Predictor", subtitle = label) +
      theme(legend.position = "none")
  }
)

grid.arrange(grobs = bd_plots)
```

## Conclusion

In this notebook, we have demonstrated how **Maximum Interpretation Decomposition (MID)** bridges the gap between predictive performance and model transparency.
By using the `{midr}` package, we successfully transformed a complex LightGBM model into a structured, additive representation.

While the surrogate model fidelity may not always be perfect, the crucial advantage lies in our ability to quantify its limitations. 
Through the **uninterpreted variation ratio**, we can directly assess the complexity of the black-box model.
If the fidelity is lower than expected, it serves as a diagnostic signal that the original model relies on high-order interactions or structural complexities that extend beyond second-order effects.

Knowing the extent of this "unexplained" variance is far more valuable than operating in the dark.
It allows actuaries to make informed decisions about whether the additional complexity of a black-box model is justified by its performance, or if a more transparent structure is preferable for regulatory and risk management purposes.

As machine learning models become increasingly prevalent in insurance pricing and reserving, tools like MID will be essential for ensuring that our "black-boxes" remain accountable, reliable, and fundamentally understood.
